\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-01-19 4-2-agg.Rmd
aggregation example WIP

In Figure \ref{fig:basic-agg}. Aggregating them into averages or other statistics would give a better picture. This can be done with cubble through a hierarchical structure, where the cluster indicator is one level above the individual station.

The `weatherdata::climate_full` data has daily climate data of `r nrow(weatherdata::climate_full)` Australia stations from 2016 to 2020.

calcluate distance matrix
As an example to illustrate here, a kmean clustering algorithm based on the distance matrix is used and the number of centres is set to 20. More complex algorithms can also be used for more complex problem, as long as a mapping from each station id to the cluster id can be constructed.

<!-- Now we can obtain the aggregated series as described in the workflow diagram above and construct the glyph map with `GGally::glyphs()`: -->

<!-- We can also look at the precipitation of each individual station within the same cluster: -->

<!-- Lastly, there is one series on Tasmania island standing out from others, lets look at where it is: -->

>>>>

With recent climate, there are `r nrow(weatherdata::climate_full)` stations in Australia that records daily maximum and minimum temperature and precipitation. While it would be too much series to show on a glyph map, stations close to each other should have similar precipitation pattern. Aggregation can be helpful here to cluster series into groups and visualise them with glyph map.

A simple kmean algorithm is used based on the distance matrix to cluster the stations into 20 groups. 
This creates `station_nested` as a station level nested cubble with a cluster column indicating which group each station belongs to. More complex algorithms can be used for more complex problem, as long as a mapping from each station id to the cluster id can be constructed.

```{r}
load(here::here("data/station_nested.rda"))
station_nested
```

To create a group level cubble, use `switch_key()` with the new key variable, `cluster`:

```{r echo = TRUE}
cluster_nested <- station_nested %>% switch_key(cluster) 
```

With the group level cubble, it is useful to compute the centroid of each cluster as the location of group level glyph map. `get_centroid()` is a shortcut for doing so or you can manually find the convex hull, its centroid, and extract the longitude and latitude:

```{r eval = FALSE, echo = TRUE}
cluster_nested <- cluster_nested %>% 
  get_centroid()
```

Long form cubble at both levels can be access through stretching the nested form. Here weekly averages of the precipitation is calculated:

```{r eval = FALSE, echo = TRUE}
station_long <- station_nested %>% 
  stretch(ts) %>% 
  mutate(wk = lubridate::week(date)) %>% 
  group_by(wk) %>% 
  summarise(prcp = sum(prcp, na.rm = TRUE))

cluster_long <- cluster_nested %>% 
  stretch(ts) %>% 
  mutate(wk = lubridate::week(date)) %>% 
  group_by(wk) %>% 
  summarise(prcp = mean(prcp, na.rm = TRUE)) %>% 
  migrate(cent_long, cent_lat)
```

With the four cubbles: `station_nested`, `station_long`, `cluster_nested`, and `cluster_long`, all the information at both levels are accessible. This allows us to make glyph map on the cluster level, looking at station composition of each cluster, and plot the individual series to check for consistency within groups. Figure \ref{fig:basic-agg} composes of the three plots together to show the precipitation in Australia.

### 2022-01-19 main.tex
aggregation example WIP

Spatial locations can have further grouping structure either in nature
(i.e.~countries are nested in continents), or calculated by some
clustering algorithms. Rather than investigate the read from individual
locations, group character summarised from multiple locations can give a
clearer picture of the local dynamic. In cubble, \texttt{switch\_key()}
can be used to create a new level of grouping of spatial location by
specifying a clustering variable. Figure \ref{fig:illu-hier} illustrates
the relationship of cubbles at site and cluster level, in both the long
and nested form. By specifying
\texttt{cluster\_nested\ \textless{}-\ station\_nested\ \%\textgreater{}\%\ switch\_key(key\ =\ cluster)},
the cubble re-defines the cubble key from \texttt{id} in
\texttt{station\_nested} to \texttt{cluster} in
\texttt{cluster\_nested}. All the spatial variables variant to
\texttt{cluster} is now nested into a \texttt{.val} column and cluster
level variables can be computed in the same fashion as station level
variables in \texttt{station\_nested}. Following the same principal of
\texttt{stretch}, the long form cluster level cubble expands the time
series variables with both indices: \texttt{cluster} and \texttt{id}.
All the cluster level variables are stored separately for recovering the
nested from from the long form.

@@

In Figure \ref{fig:basic-agg}. Aggregating them into averages or other
statistics would give a better picture. This can be done with cubble
through a hierarchical structure, where the cluster indicator is one
level above the individual station.

The \texttt{weatherdata::climate\_full} data has daily climate data of
639 Australia stations from 2016 to 2020.

calcluate distance matrix As an example to illustrate here, a kmean
clustering algorithm based on the distance matrix is used and the number
of centres is set to 20. More complex algorithms can also be used for
more complex problem, as long as a mapping from each station id to the
cluster id can be constructed.

>>>>

Spatial locations can have further grouping structure either in nature
(i.e.~countries are nested in continents), or calculated by some
clustering algorithms. Rather than investigate the read from individual
locations, group character summarised from multiple locations can give a
clearer picture of the local dynamic. In cubble, \texttt{switch\_key()}
can be used to create a new level of grouping of spatial location by
specifying a clustering variable. Figure \ref{fig:illu-hier} illustrates
the relationship of cubbles at site and cluster level, in both the long
and nested form. By specifying
\texttt{cluster\_nested\ \textless{}-\ station\_nested\ \%\textgreater{}\%\ switch\_key(key\ =\ cluster)},
the cubble re-defines the cubble key from \texttt{id} in
\texttt{station\_nested} to \texttt{cluster} in
\texttt{cluster\_nested}. All the spatial variables variant to
\texttt{cluster} are now nested into a \texttt{.val} column and cluster
level variables can be computed in the same fashion as station level
variables in \texttt{station\_nested}. Following the same principal of
\texttt{stretch}, the long form cluster level cubble expands the time
series variables with both indices: \texttt{cluster} and \texttt{id}.
All the cluster level variables are stored separately for recovering the
nested from from the long form.

@@

With recent climate, there are 639 stations in Australia that records
daily maximum and minimum temperature and precipitation. While it would
be too much series to show on a glyph map, stations close to each other
should have similar precipitation pattern. Aggregation can be helpful
here to cluster series into groups and visualise them with glyph map.

A simple kmean algorithm is used based on the distance matrix to cluster
the stations into 20 groups. This creates \texttt{station\_nested} as a
station level nested cubble with a cluster column indicating which group
each station belongs to. More complex algorithms can be used for more
complex problem, as long as a mapping from each station id to the
cluster id can be constructed.

\begin{CodeChunk}
\begin{CodeOutput}
# cubble:   id [639]: nested form
# bbox:     [113.53, -43.66, 153.64, -10.05]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name             wmo_id ts               cluster
  <chr>       <dbl> <dbl> <dbl> <chr>             <dbl> <list>             <int>
1 ASN00001006 -15.5  128.   3.8 wyndham aero      95214 <tibble [1,827 ~       3
2 ASN00001007 -13.8  126.   6   troughton island  94102 <tibble [1,827 ~       3
3 ASN00001018 -16.4  126. 546   mount elizabeth   94211 <tibble [1,460 ~      16
4 ASN00001019 -14.3  127.  23   kalumburu         94100 <tibble [1,827 ~       3
5 ASN00001020 -14.1  126.  51   truscott          95101 <tibble [1,827 ~       3
# ... with 634 more rows
\end{CodeOutput}
\end{CodeChunk}

To create a group level cubble, use \texttt{switch\_key()} with the new
key variable, \texttt{cluster}:

\begin{CodeChunk}
\begin{CodeInput}
R> cluster_nested <- station_nested %>% switch_key(cluster) 
\end{CodeInput}
\end{CodeChunk}

With the group level cubble, it is useful to compute the centroid of
each cluster as the location of group level glyph map.
\texttt{get\_centroid()} is a shortcut for doing so or you can manually
find the convex hull, its centroid, and extract the longitude and
latitude:

\begin{CodeChunk}
\begin{CodeInput}
R> cluster_nested <- cluster_nested %>% 
+   get_centroid()
\end{CodeInput}
\end{CodeChunk}

Long form cubble at both levels can be access through stretching the
nested form. Here weekly averages of the precipitation is calculated:

\begin{CodeChunk}
\begin{CodeInput}
R> station_long <- station_nested %>% 
+   stretch(ts) %>% 
+   mutate(wk = lubridate::week(date)) %>% 
+   group_by(wk) %>% 
+   summarise(prcp = sum(prcp, na.rm = TRUE))
R> 
R> cluster_long <- cluster_nested %>% 
+   stretch(ts) %>% 
+   mutate(wk = lubridate::week(date)) %>% 
+   group_by(wk) %>% 
+   summarise(prcp = mean(prcp, na.rm = TRUE)) %>% 
+   migrate(cent_long, cent_lat)
\end{CodeInput}
\end{CodeChunk}

With the four cubbles: \texttt{station\_nested}, \texttt{station\_long},
\texttt{cluster\_nested}, and \texttt{cluster\_long}, all the
information at both levels are accessible. This allows us to make glyph
map on the cluster level, looking at station composition of each
cluster, and plot the individual series to check for consistency within
groups. Figure \ref{fig:basic-agg} composes of the three plots together
to show the precipitation in Australia.
